data preparation

### 2024-07-05 data_preparation.Rmd
data preparation

---
title: "manuscript_analysis"
author: "Sophie Moreau"
date: "2024-07-03"
output: html_document
editor_options: 
  chunk_output_type: console
---

@@

#2. Data Analysis
##General Additive Models
Running the predictions for all GAMs with the smaller data frames to reduce running 
time.

```{r}
# packages relevant for data analysis
library(tidyverse)
library(here)
library(readr)
library(readxl)
library(gratia)
library(mgcv)
library(forcats)
library(ggpubr)
library(RColorBrewer)
library(broom)
library(grid)
library(DHARMa)
library(mgcViz)
library(gridExtra)
library(grid)

# function for the GAMs (old)
# source(here("function_GAM.R"))

# new function for temp and depth models 
source(here("function_temp_depth.R"))
```

###Predictions of GAMs

```{r}
# read in the dataframes
df_1 <- readRDS("data_frame_models/df_binomial_gam")
df_2 <- readRDS("data_frame_models/df_abundance_gam")
df_3 <- readRDS("data_frame_models/df_binomial_re")
df_4 <- readRDS("data_frame_models/df_abundance_re")

# use own function predictions() from function_GAM.R to predict species' models
# across all lakes

# predictions take VERY LONG! and two folders are needed to store the 
# DHARMa and prediction output inside: 
# total_models/gam_check
# total_models/predictions

# DO NOT RUN NOW, RUN LATER!
predictions_depth_temp(df_1)
predictions_depth_temp(df_2)
predictions_depth_temp(df_3)
predictions_depth_temp(df_4)

# issue with cottus sp po
# statistics of the gamms
total_models <- bind_rows(df_1, df_2, df_3, df_4)

stats_gamms(total_models)

#  compile data frame with all predictions from the saved species' model predictions

df_predictions_all <- list.files(path = "total_models/predictions", pattern = ".rds", full.names = TRUE) |>
  map_dfr(readRDS)

# save total predictions as RDS
saveRDS(df_predictions_all, "total_models/df_pred_all.rds")

###AIC comparisons in special cases

```{r}

AIC_testing <- function(data2, i){

  data <- data2 |> 
    filter(Species == i)

  data$fProtocol <- as.factor(data$Protocol)
  data$fLake <- as.factor(data$Lake)

  # model_0 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) +
  #                  s(Depth_sample, k = 3)
  #                           + s(fLake, bs = 're')
  #                           +  s(fProtocol, bs = 're'), family = ziP())
  # print(glance(model_0))

model1 <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) + s(Depth_sample, k = 3)
                            + s(fLake, bs = 're')
                            +  s(fProtocol, bs = 're'), family = binomial)
print(glance(model1))

model2 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) +
                            + s(fLake, bs = 're')
                            +  s(fProtocol, bs = 're'), family = ziP())
print(glance(model2))

}

AIC_testing(df_4, "Lepomis_gibbosus")

# in Alosa agone: model_0 not running, smaller AIC for binomial depth + temp model -> will be modelled as binomial depth + temp
# same for Coregonus_sp_large_pelagic
# lepomis gibbosus also has the smallest AIC for binomial depth + temp model

# also testing the other special cases:
 # "Alburnus_arborella", "Barbatula_sp_Lineage_I","Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta"

AIC_testing(df_4, "Alburnus_arborella") 
# not running with model0, Binomial with depth and temp best
AIC_testing(df_4, "Barbatula_sp_Lineage_I")
# binomial depth and temp has lowest AIC
AIC_testing(df_4, "Cyprinus_carpio")
# binomial depth and temp has lowest AIC 
AIC_testing(df_4, "Phoxinus_csikii")
#binomial depth and temp has lowest AIC 
AIC_testing(df_4, "Salmo_trutta")
# binomial depth and temp lowest AIC

# AIC_testing(df_4, "Cottus_sp_Po")
# # best model is the temp, depth model

AIC_testing(total_models, "Alburnus_arborella")
# 8 models are run with binomial temp and depth

```
####Fig. 2
Overview of all model predictions

```{r, fig.show='hide'}
# Load total predictions
model_predictions <- readRDS("total_models/df_pred_all.rds")

# check the species list
model_predictions$species <- as.factor(model_predictions$species)
levels(model_predictions$species)

# read in the category table and rename categories
species_across_lakes <-  read_xlsx("species_category_across_lakes.xlsx") |> 
  select(-notes) |> 
    mutate(category = ifelse(category == "native", "non_endemic_native", category)) |>
  mutate(category = ifelse(category == "non_native_region", "non_endemic_native_and_translocated", category)) |> 
  mutate(category2 = factor("all"))

species_across_lakes$category <- as.factor(species_across_lakes$category)
levels(species_across_lakes$category)

#merge predictions with categories
predictions_categories <- merge(model_predictions, species_across_lakes)

# EVTL NEW PLOT 2
# change category names for facet_wrap
category_names <- c(
  `endemic` = "endemic",
  `endemic_and_translocated` = "translocated endemic",
  `non_endemic_native` = "non-endemic native",
  `non_endemic_native_and_translocated` = "translocated non-endemic native",
  `non_native` = "non-native"
)

# colors for the plot
mycolors1 <-  c("endemic"= "#63D8F2", "non_endemic_native"="#F26379", "non_native" = "#5A64D8",
                "endemic_and_translocated" = "#F2AD63", "non_endemic_native_and_translocated" = "#F2AD63")

# Rescale the predictions to 0-1 
predictions_categories_rescaled <- predictions_categories |>
  group_by(species) |>
  mutate(fit_rescaled = (fit - min(fit)) / (max(fit) - min(fit)))

plot_pred <- predictions_categories_rescaled |>
  ggplot(aes(temp, fit_rescaled, group = species, color = category)) +
  geom_line() +
  # how can I get the right color?
   # geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit)), alpha = 0.3) +
  theme_bw(base_size = 16) +
  ylab("Abundance") +
  xlab("Temperature") +
  scale_color_manual(values = mycolors1, guide = NULL) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        legend.position = "bottom",
        legend.title = element_blank())

plot_category_predictions <- plot_pred + facet_grid(~category, labeller = as_labeller(category_names))

plot_category_predictions

# trying to group by category and color
figure_test_2 <- predictions_categories |>
  ggplot(aes(temp, fit)) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit), fill = factor(category)), alpha = 0.3) +
  geom_line(aes(color = factor(category))) +
  theme_bw() +
  facet_wrap(~species + category, scale = "free_y", labeller = labeller(species = label_wrap_gen(width = 10))) +
  theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
  scale_color_manual(values = mycolors1, aesthetics = c("color", "fill")) +
  xlab("Temperature") +
  ylab("Abundance")

# tiff(paste("total_models/plots/draft_all_predictions.tiff", sep = ""), units="in", width = 12, height=8, res=300)
# plot(figure_test_2)
# 
# # Closing the graphical device
# dev.off()

# OLD FIGURE 2

all_predictions <- predictions_categories |>
  ggplot(aes(temp, fit)) +
  geom_ribbon(aes(ymin = (fit - se.fit), ymax = (fit + se.fit), fill = factor(species)), alpha = 0.3) +
  geom_line(aes(color = factor(species))) +
  theme_bw() +
  facet_wrap(~species, scale = "free_y", labeller = labeller(species = label_wrap_gen(width = 10))) +
  theme(strip.background = element_rect(fill = "lightgrey"), strip.text = element_text(size = 6)) +
  scale_color_viridis(discrete = TRUE, guide = NULL, option = "G", aesthetics = c("color", "fill")) +
  xlab("Temperature") +
  ylab("Abundance")

all_predictions
# save the plot as TIFF
# tiff(paste("total_models/plots/all_predictions_models.tiff", sep = ""), units="in", width = 12, height=8, res=300)
# plot(all_predictions)
# 
# # Closing the graphical device
# dev.off()

##Sensitivity testing
NOT CHANGED 
Testing if depth-driven temperature models can adequately fit abundance data compared to
a temperature and depth model. 
Models were compared in the eleven most abundant species.

```{r}
# Read in abundance and depth data of all species included for modelling
df_models <- readRDS("df_models.rds")

# find 11 most abundant  species
species_list <- df_models |> 
  filter(tot_abu > 300) |> 
  distinct(Species) |> 
  pull(Species)

# most abundant species
 # "Alburnus_alburnus", "Coregonus_sp_albeli", "Coregonus_sp","Gasterosteus_aculeatus",
 # "Gobio_gobio", "Leuciscus_leuciscus", "Perca_fluviatilis", "Rutilus_rutilus", 
 # "Sander_lucioperca", "Scardinius_erythrophthalmus","Gymnocephalus_cernua"

# filter df into the most abundant species and select the columns of interest

df_sensitivity <- df_models |> 
  filter(tot_abu > 300) |> 
  select(Lake, Species, Protocol, Depth_sample, Abundance, Presence, mean_last_7days) |> 
  #there are some NAs in depth -> drop them
  drop_na(Depth_sample, mean_last_7days)

df_sensitivity$fLake <- as.factor(df_sensitivity$Lake)
df_sensitivity$fProtocol <- as.factor(df_sensitivity$Protocol)

# Loop through the eleven species and fit temp and temp-depth model

species_list <- df_sensitivity |> 
  distinct(Species) |> 
  pull(Species)

# loop through the species, model both GAMs, check DHARMa and predictions
for (i in species_list){

  data <- df_sensitivity |> 
    filter(Species == i)

  #Model fit for temp and depth
  model1 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(Depth_sample, k = 3)
                + s(fLake, bs = 're')
                +  s(fProtocol, bs = 're'), family = ziP())
  # Model fit for temp
  model2 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3)
                + s(fLake, bs = 're')
                +  s(fProtocol, bs = 're'), family = ziP())
  # validate model predictions for temp and depth model with dharma:
  # prepare residuals
  simulationOutput <- simulateResiduals(fittedModel = model1, plot = F)
  # tiff_filename <- paste("total_models/gam_check/temp_depth_check_", i, ".tiff", sep = "")
  # tiff(tiff_filename, width = 800, height = 600)
  print(plot(simulationOutput))
  # dev.off()
  # Plotting standardized residuals against predictors
  # tiff_file_2 <- paste("total_models/gam_check/temp_depth_predictor_", i, ".tiff", sep = "")
  # tiff(tiff_file_2, width = 800, height = 600)
  print(plotResiduals(model1, data$mean_last_7days, xlab = "temp", main=NULL))
  # dev.off()

  # for predictions with random effects in the GAMs, we select a random value
  # of the level of random effects and exclude the random effects in the prediction

  # lakes where the species is present
  unique_lakes <- distinct(data, Lake) |>
    pull()

  # select random lake for prediction
  random_lake <- sample(unique_lakes, 1)

  # for temp-depth model only predict for depth where the species is present
  presence_depth <- data |> 
    filter(Presence == 1) |>
    distinct(Depth_sample) |> 
    pull(Depth_sample)

  # take randomly 5 depth values for prediction
  random_depth <- sort(sample(presence_depth, 5))

  # Convert sorted values to factors with original order
  factor_random_depth <- factor(random_depth, levels = random_depth, ordered = TRUE)

  # grid for each of the models

  # temp-depth model predicted with random depths where the species is present
 grid1 <- expand.grid(mean_last_7days = seq(
    from = min(data$mean_last_7days, na.rm = TRUE),
    to = max(data$mean_last_7days, na.rm = TRUE), length = 200),
    fProtocol = factor("VERT"), fLake = factor(random_lake), 
    Depth_sample = factor_random_depth)

  grid2 <- expand.grid(mean_last_7days = seq(
    from = min(data$mean_last_7days, na.rm = TRUE),
    to = max(data$mean_last_7days, na.rm = TRUE), length = 1000),
    fProtocol = factor("VERT"), fLake = factor(random_lake))

  # print relevant stats
  print(paste(i))
  print(glance(model1))
  print(glance(model2))

  # manually predict both models
  model_prediction1 <- predict.gam(model1, newdata = grid1,
                                  exclude = c("s(fProtocol)", "s(fLake)"), #exclude the re
                                  term = "s(mean_last_7days)", type = "response",             
                                  se.fit = TRUE)

  model_bind1 <- cbind(grid1, as.data.frame(model_prediction1)) |>
    mutate(model = factor("temp_and_depth"))

# predict model2
  model_prediction2 <- predict.gam(model2, newdata = grid2,
                                   exclude = c("s(fProtocol)", "s(fLake)"),
                                   type = "response", se.fit = TRUE)

  model_bind2 <- cbind(grid2, as.data.frame(model_prediction2)) |>
    mutate(model = factor("temp"))

  df_pred <- bind_rows(model_bind1, model_bind2) |>
    rename(temp = mean_last_7days) |>
    select(-fProtocol, -fLake) |>
    mutate(Depth_sample = factor(ifelse(is.na(Depth_sample), "temp", as.character(Depth_sample))))

  # Extract numeric depth levels
  numeric_levels <- levels(df_pred$Depth_sample)[!levels(df_pred$Depth_sample) %in% "temp"]

  # Sort numeric depth levels
  sorted_numeric_levels <- sort(as.numeric(numeric_levels))

  # Combine sorted numeric levels with "temp" level
  sorted_levels <- c(as.character(sorted_numeric_levels), "temp")

  # Reorder levels
  df_pred$Depth_sample <- factor(df_pred$Depth_sample, levels = sorted_levels)

  # Check the levels to confirm they are ordered from lowest to highest
  levels(df_pred$Depth_sample)

  # rescale predictions of both models
  df_pred_rescaled <- df_pred |>
    group_by(model) |>
    mutate(fit_rescaled = (fit - min(fit)) / (max(fit) - min(fit))) |>
    mutate(se_fit_rescaled = (se.fit - min(se.fit)) / (max(se.fit) - min(se.fit))) |> 
    mutate(species = factor(i))

  # save predictions of both models in a rds
   saveRDS(df_pred_rescaled, paste0("total_models/sensitivity/rds/sensitivity_predictions_", i, ".rds"))

}

# prepare total predictions of all models
sensitivity_predictions <- list.files(path = "total_models/sensitivity/rds", pattern = ".rds", full.names = TRUE) |>
  map_dfr(readRDS)

# save total predictions as RDS
saveRDS(sensitivity_predictions, "total_models/sensitivity_predictions.rds")

###Fig. S1

```{r}
# read in the sensitivity testing data
sensitivity_predictions <- readRDS("total_models/sensitivity_predictions.rds")

levels(sensitivity_predictions$species)

# Plot both model predictions together per species
 figure_s1 <- sensitivity_predictions |>
    ggplot(aes(temp, fit_rescaled, color = model, fill = model)) +
    geom_line() +
    geom_ribbon(aes(ymin = (fit_rescaled - se.fit), ymax = (fit_rescaled + se.fit)), color = NA, alpha = 0.1) +
    theme_bw() +
    # labs(title = paste("Sensitivity Test", i)) +
    scale_color_manual(values = c("#08306B", "#D73027"), aesthetics = c("color", "fill")) +
    facet_wrap(~species, scale = "free") +
    ylab("Abundance") +
    xlab("Temperature") +
   theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        legend.position = "right",
        legend.title = element_blank())

  figure_s1
  # save as TIFF
# tiff(paste("total_models/plots/sensitivity_overview.tiff", sep = ""), units="in", width=11, height=7, res=300)
# 
# plot(figure_s1)
# 
# dev.off()

```
##Derivatives calculation

We calculated derivatives for each species per lake.

```{r}

# Derivatives() will ignore any random effect smooths it encounters in object.

# Table of pecies that were recorded with electro, fishbase or eawag, following Table 16 in Projet Lac Synthesis Report. 
# Plus the minimum and maximum temp in each lake -> needed for the new data for 
# calculating derivatives

species_lake <- read_xlsx("species_occurrences_lakes.xlsx") 
str(species_lake)

species_lake$fLake <- as.factor(species_lake$Lake)
species_lake$fProtocol <- as.factor(species_lake$Protocol)
species_lake$Species <- as.factor(species_lake$Species)

#Loading all four dfs to calculate derivatives

df_binomial_gam <- readRDS("data_frame_models/df_binomial_gam")
df_binomial_gam$fLake <- as.factor(df_binomial_gam$Lake)
df_binomial_gam$fProtocol <- as.factor(df_binomial_gam$Protocol)

df_abundance_gam <- readRDS("data_frame_models/df_abundance_gam")
df_abundance_gam$fProtocol <- as.factor(df_abundance_gam$Protocol)
df_abundance_gam$fLake <- as.factor(df_abundance_gam$Lake)

df_binomial_re <- readRDS("data_frame_models/df_binomial_re")
df_binomial_re$fLake <- as.factor(df_binomial_re$Lake)
df_binomial_re$fProtocol <- as.factor(df_binomial_re$Protocol)

df_abundance_re <- readRDS("data_frame_models/df_abundance_re")
df_abundance_re$fLake <- as.factor(df_abundance_re$Lake)
df_abundance_re$fProtocol <- as.factor(df_abundance_re$Protocol)

# separating the dfs to separate model types
depth_binomial_gam <- df_binomial_gam |> 
  filter(tot_abu > 20)

temp_binomial_gam <- df_binomial_gam |> 
  filter(tot_abu < 20)

depth_abundance_gam <- df_abundance_gam |> 
  filter(tot_abu > 20)

temp_abundance_gam <- df_abundance_gam |> 
  filter(tot_abu < 20)

depth_binomial_re <- df_binomial_re |> 
  filter(tot_abu > 20)

temp_binomial_re <- df_binomial_re |> 
  filter(tot_abu < 20)

depth_abundance_re <- df_abundance_re |> 
  filter(tot_abu > 20)

temp_abundance_re <- df_abundance_re |> 
  filter(tot_abu < 20)

###Depth and temp models
```{r}
# DEPTH AND TEMPERATURE
# UNCLEAR HOW THIS CAN BE DONE
# Model 1: Binomial data without random intercepts

species_list <- depth_binomial_gam |>
  distinct(Species) |>
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

i <- "Salmo_sp_Blackspot"
j <- "Poschiavo"
# loop through the list of species
for (i in species_list) {
  data <- depth_binomial_gam |>
    filter(Species == i)

  lake_data <- species_lake |>
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) +
                           s(Depth_sample, k = 3) + s(fProtocol, bs = 're'),
                         family = binomial)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  # loop through all lakes where the species is present, lake list is based on the information from
  # Projet Lac
  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    # using minimum and maximum temp in the lake for derivatives
     newdata <- tibble(mean_last_7days = seq(from = min(data_lake$temp, na.rm = TRUE),
                                             to = max(data_lake$temp, na.rm = TRUE)
                                             ,length = 200), Depth_sample =
                         mean(data$Depth_sample, na.rm = TRUE),
                       fProtocol = factor("VERT"))

     # calculate derivatives
     # derivatives <- derivatives(gam_output[[i]], data = newdata) |>
     #   mutate(fLake = factor(j)) |>
     #   mutate(species = factor(i))
     #   rename(temp = data)
     #

       test_derivatives <- derivative_samples(gam_output[[i]]
                                              , focal = "mean_last_7days", data = newdata)

       test_derivatives |>
         filter(.draw <= 200) |>
         ggplot(aes(x = mean_last_7days, y = .derivative, group = .draw)) +
         geom_line(alpha = 0.5)

       derivatives <- test_derivatives |> 
         filter(.draw == 500) |> 
         mutate(fLake = factor(j)) |>
         mutate(species = factor(i))

       derivatives |> 
         ggplot(aes(mean_last_7days, .derivative)) +
         geom_line()

     saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
  }
}

# needs to be discussed

# nothing changed starting from here
############################################################################3##3
# Model 2.1: Abundance data and no random intercept for lake -> but binomial
# Coregonus_sp_benthic_profundal did not run with ZIP and was thus
# run with binomial.

species_list <- depth_abundance_gam |> 
  # binomial one
  filter(Species == "Coregonus_sp_benthic_profundal") |>
  distinct(Species) |> 
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {

  data <- df_abundance_gam |> 
    filter(Species == i)

  # binomial family instead of ZIP
  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) +
                           s(fProtocol, bs = 're'), family = binomial())

  lake_data <- species_lake |>
    filter(Species == i)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), length = 200),
      fProtocol = factor("VERT"))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = data)

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
  }
}

# Model 2.2: Abundance data and no random intercept for lake -> ZIP
# all other species in this group run with ZIP

species_list <- df_abundance_gam |>
  filter(!Species == "Coregonus_sp_benthic_profundal") |>
  distinct(Species) |> 
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {

  data <- df_abundance_gam |> 
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fProtocol, bs = 're'),
                         family = ziP())

  lake_data <- species_lake |>
    filter(Species == i)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), length = 200),
      fProtocol = factor("VERT"))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = data)

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
  }
}

# Model 3: Binomial data and random intercept for lake

species_list <- df_binomial_re |> 
  distinct(Species) |> 
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {

  data <- df_binomial_re |> 
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = "re")
                         + s(fProtocol, bs = 're'), family = binomial)

  lake_data <- species_lake |>
    filter(Species == i)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  unique_lakes <- distinct(data, Lake) |> 
    pull()

  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    # take all lakes where the species is present and take random one for 
    # the data included in derivatives()
    random_lake <- sample(unique_lakes, 1)

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), lenqh = 200),
      fProtocol = factor("VERT"), fLake = factor(random_lake))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = data)

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))

  }
}

# Model 4.1: Abundance data and random intercept for lake, simplified models
# multiple species in this group were simplified and run with binomial

species_list <- df_abundance_re |>
  # simplified models
  filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
                                "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta",
                        "Lepomis_gibbosus")) |>
  distinct(Species) |>
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {

  data <- df_abundance_re |>
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
                         +  s(fProtocol, bs = 're'), family = binomial)

  lake_data <- species_lake |>
    filter(Species == i)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  unique_lakes <- distinct(data, Lake) |>
    pull()

  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    random_lake <- sample(unique_lakes, 1)

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), length = 200),
      fLake = factor(random_lake), fProtocol = factor("VERT"))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = data)
    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))

  }
}

# Model 4.1: Abundance data and random intercept for lake, complex models

species_list <- df_abundance_re |> 
  # exclusion of simplified models
  filter(!Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
                        "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta", 
                        "Lepomis_gibbosus")) |>
  distinct(Species) |> 
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {

  data <- df_abundance_re |> 
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
                         +  s(fProtocol, bs = 're'), family = ziP())

  lake_data <- species_lake |>
    filter(Species == i)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  unique_lakes <- distinct(data, Lake) |> 
    pull()

  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    random_lake <- sample(unique_lakes, 1)

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), length = 200),
      fLake = factor(random_lake), fProtocol = factor("VERT"))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = data)

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))

  }
}

# save all the derivatives in one file

df_deriv <- list.files(path = "total_models/derivatives", pattern = ".rds", full.names = TRUE) |> 
  map_dfr(readRDS)

saveRDS(df_deriv, "total_models/df_deriv_all.rds")

# save total derivatives as RDS (no exclusions)
# saveRDS(df_deriv, "total_models/df_deriv_all_no_excl.rds")

###Temp models
```{r}

# Now four different loops are run to calculate derivatives of each species 
# in each lake

# Model 1: Binomial data without random intercepts

species_list <- df_binomial_gam |>
  distinct(Species) |>
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

# loop through the list of species
for (i in species_list) {
  data <- df_binomial_gam |>
    filter(Species == i)

  lake_data <- species_lake |>
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fProtocol, bs = 're'), family = binomial)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  # loop through all lakes where the species is present, lake list is based on the information from
  # Projet Lac
  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    # using minimum and maximum temp in the lake for derivatives
     newdata <- tibble(mean_last_7days = seq(from = min(data_lake$temp, na.rm = TRUE),
                                             to = max(data_lake$temp, na.rm = TRUE), length = 200),
                       fProtocol = factor("VERT"))

     # calculate derivatives
     derivatives <- derivatives(gam_output[[i]], data = newdata) |>
       mutate(fLake = factor(j)) |>
       mutate(species = factor(i)) |>
       rename(temp = data)

     saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
  }
}

# Model 2.1: Abundance data and no random intercept for lake -> but binomial
# Coregonus_sp_benthic_profundal did not run with ZIP and was thus
# run with binomial.

species_list <- df_abundance_gam |> 
  # binomial one
  filter(Species == "Coregonus_sp_benthic_profundal") |>
  distinct(Species) |> 
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {

  data <- df_abundance_gam |> 
    filter(Species == i)

  # binomial family instead of ZIP
  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) +
                           s(fProtocol, bs = 're'), family = binomial())

  lake_data <- species_lake |>
    filter(Species == i)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), length = 200),
      fProtocol = factor("VERT"))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = data)

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
  }
}

# Model 2.2: Abundance data and no random intercept for lake -> ZIP
# all other species in this group run with ZIP

species_list <- df_abundance_gam |>
  filter(!Species == "Coregonus_sp_benthic_profundal") |>
  distinct(Species) |> 
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {

  data <- df_abundance_gam |> 
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fProtocol, bs = 're'),
                         family = ziP())

  lake_data <- species_lake |>
    filter(Species == i)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), length = 200),
      fProtocol = factor("VERT"))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = data)

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
  }
}

# Model 3: Binomial data and random intercept for lake

species_list <- df_binomial_re |> 
  distinct(Species) |> 
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {

  data <- df_binomial_re |> 
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) + s(fLake, bs = "re")
                         + s(fProtocol, bs = 're'), family = binomial)

  lake_data <- species_lake |>
    filter(Species == i)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  unique_lakes <- distinct(data, Lake) |> 
    pull()

  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    # take all lakes where the species is present and take random one for 
    # the data included in derivatives()
    random_lake <- sample(unique_lakes, 1)

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), lenqh = 200),
      fProtocol = factor("VERT"), fLake = factor(random_lake))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = data)

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))

  }
}

# Model 4.1: Abundance data and random intercept for lake, simplified models
# multiple species in this group were simplified and run with binomial

species_list <- df_abundance_re |>
  # simplified models
  filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
                                "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta",
                        "Lepomis_gibbosus")) |>
  distinct(Species) |>
  pull(Species)

species_list <- sort(species_list)

derivatives <- list()
gam_output <- list()
model_prediction <- list()

for (i in species_list) {

  data <- df_abundance_re |>
    filter(Species == i)

  gam_output[[i]] <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) + s(fLake, bs = 're')
                         +  s(fProtocol, bs = 're'), family = binomial)

  lake_data <- species_lake |>
    filter(Species == i)

  lake_list <- distinct(lake_data, Lake) |>
    pull()

  unique_lakes <- distinct(data, Lake) |>
    pull()

  for (j in lake_list){

    data_lake <- species_lake |>
      filter(Species == i) |>
      filter(Lake == j)

    random_lake <- sample(unique_lakes, 1)

    newdata <- tibble(mean_last_7days = seq(
      from = min(data_lake$temp, na.rm = TRUE),
      to = max(data_lake$temp, na.rm = TRUE), length = 200),
      fLake = factor(random_lake), fProtocol = factor("VERT"))

>>>>

---
title: "data preparation"
author: "Sophie Moreau"
date: "2024-07-05"
output: html_document
editor_options: 
  chunk_output_type: console
---

@@
